---
output:
  html_document: default
  pdf_document: default
---
Activity Monitoring Analysis
========================================================
Melanie Butler

Peer Assessment 1 for Reproducible Research

## Get the Data
In this section, we download the data.

This code chunk downloads the .CSV file with the data, unzips it, and appropriately changes the working directory again.
```{r downloaddata, echo=TRUE}
##Dowload data, unzip, and set working directory
fileURL<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileURL, "ActivityData.zip")
unzip("ActivityData.zip", exdir = "ActivityData")
```

This next chunck reads in the data and prints a summary.
```{r ReadData, echo=TRUE}
AD<-read.csv("activity.csv")
summary(AD)
```

## Process the Data

In this section we process the data so that we can begin the analysis. We also load libraries that we need.

In this first chunk in this section, we coerce the date column from a factor variable to a date variable.
```{r DateVariable, echo=TRUE}
AD$date<-as.Date(AD$date)
```

We load ggplot2, dplyr, plyr, and gridExtra.
```{r loadlibraries, echo=TRUE}
library(ggplot2)
library(dplyr)
library(plyr)
library(gridExtra)
```

## Total Number of Steps Each Day

In this section we want to get a general feel for the data. We ignore missing values, which are reported in the data as NA.

In this first code chunk, we make a histogram of the total number of steps taken each day.
```{r TotalStepsEachDay, echo=TRUE}
dailytotal<-aggregate(AD$steps, FUN=sum, by=list(Group.date=AD$date), na.rm=TRUE)
hist(dailytotal$x, col="green", breaks = 25, xlab = "Number of Steps", 
     ylab ="Frequency", main = "Total Number of Steps Taken Each Day" )
```

In this second code chunk we calulate the mean and median number of steps taken each day.
```{r meanandmedian, echo=TRUE}
M<-mean(dailytotal$x)
Med<-median(dailytotal$x)
```
The mean is `r M`. The median is `r Med`.

## 5-minute intervals
In this section we explore the data grouped by the five-minute intervals, still ignoring the missing (NA) values.


In this first code chunk, we make a time series plot of the average number of steps taken in each five-minute interval.
```{r timeseriesplot, echo=TRUE}
intervals<-aggregate(AD$steps, FUN=mean,
                     by=list(Group.Interval=AD$interval), na.rm=TRUE)
ggplot(intervals, aes(Group.Interval, x))+
    geom_line() +
    xlab("Interval") + ylab("Average Steps")
```

In this next code chunk, we find the five-minute interval that on average contains the maximum number of steps.
```{r maxstepsinterval}
maximum<-max(intervals$x)
maxstepinterval<-filter(intervals, x>206)
maxstepinterval<-maxstepinterval[1,1]
```
The time interval that on average contains the maximum number of steps is `r maxstepinterval`.

## Imputing Missing Data
In this section, we impute the missing data. Without knowing more about how the data was collected, we do not know why the data is missing. For this reason, it makes the most sense to use the mean number of steps for a given day to replace any of the NA values for that day.

In this first code chunk, we calculate the total number of missing (NA) data values.
```{r numberofNA}
missing<-sum(is.na(AD$steps))
```
The total number of rows with missing data values is `r missing`.

In this next code chunk we impute the missing data values by replacing missing data values with the mean number of steps for that day. In the case of missing data over every interval in a day, we remove that day from the analysis.
```{r imputingmissingdata, echo=TRUE}
AD2<-read.csv("activity.csv")
AD2$date<-as.character(AD2$date)
AD2$date<-factor(AD2$date, labels=c(1:61))
imputedmean<-NULL
for (i in 1:61){
    imputedmean<-c(imputedmean,
        ifelse(is.na(AD2[which(AD2$date==i),]$steps), 
               mean(AD2[which(AD2$date==i),]$steps, na.rm= TRUE), 
               AD2[which(AD2$date==i),]$steps))
}
AD3<-cbind(AD2[,2:3], imputedmean)
AD3<-na.omit(AD3)
```

In this code chunk, we make a histogram of the total number of steps taken each day after we have imputed the missing data values.
```{r TotalStepsEachDayImputed, echo=TRUE}
dailytotal2<-aggregate(AD3$imputedmean, FUN=sum, by=list(Group.date=AD3$date), na.rm=TRUE)
hist(dailytotal2$x, col="green", breaks = 25, xlab = "Number of Steps", 
     ylab ="Frequency", main = "Total Number of Steps Taken Each Day" )
```

Note that there does not appear to be a difference in the two histograms (total number of steps taken each day before and after imputing), except for the number of days with 0 steps.

## Weekdays vs Weekends
In this section we compare the number of steps taken per 5-minute interval in weekdays vs weekends.

In this code chunk, we separate the data into two dataframes - one for weekdays and one for Weekends.
``` {r weekdayvsweekend, echo=TRUE}
AD4<-cbind(AD, weekdays(AD$date))
names(AD4)<-c("steps", "date", "interval", "dayofweek")
Monday<-AD4[which(AD4$dayofweek=="Monday"),]
Tuesday<-AD4[which(AD4$dayofweek=="Tuesday"),]
Wednesday<-AD4[which(AD4$dayofweek=="Wednesday"),]
Thursday<-AD4[which(AD4$dayofweek=="Thursday"),]
Friday<-AD4[which(AD4$dayofweek=="Friday"),]
Saturday<-AD4[which(AD4$dayofweek=="Saturday"),]
Sunday<-AD4[which(AD4$dayofweek=="Sunday"),]
MTWHF<-join(join(join(join(Monday, Tuesday), Wednesday), Thursday), Friday)
SatSun<-join(Saturday, Sunday)
```

In this code chunk, we make a two time series plots of the average number of steps taken in each five-minute interval on weekdays versus weekends.
```{r timeseriesplot2, echo=TRUE}
intervalsweekday<-aggregate(MTWHF$steps, FUN=mean,
                     by=list(Group.Interval=MTWHF$interval), na.rm=TRUE)
intervalsweekend<-aggregate(SatSun$steps, FUN=mean,
                     by=list(Group.Interval=SatSun$interval), na.rm=TRUE)
p1<-ggplot(intervalsweekday, aes(Group.Interval, x))+
    geom_line() +
    xlab("Weekdays") + ylab("Average Steps")
p2<-ggplot(intervalsweekend, aes(Group.Interval, x))+
    geom_line() + xlab("Weekends") + ylab("Average Steps")
grid.arrange(p1, p2, nrow=2)
```

From these plots, we see a later start to steps on the weekends (probably the person is sleeping in on the weekends) and more steps in the middle of the day on weekends (probably because the person is not working at a desk).